Experience Report: Haskell in Computational Biology

### 2012-06-28 mrfy_experience_report.tex
ccs stuff added
new explanation of function vee-prime

\def\goo{18pt}
\def\gum{14pt}
\[
\begin{array}{l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
\end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
\vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
\end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
\vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
\end{array} \right.\\
\end{array}
\qquad
\begin{array}{l@{}c@{}l}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{\prime M}(i - 1)} {a^{\prime}_{M_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{\prime I}(i - 1)} {a^{\prime}_{I_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{\prime D}(i - 1)} {a^{\prime}_{D_{j-1}M_{j}}}\\
\end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j}^{\prime M}(i - 1)} {a^{\prime}_{M_{j}I_{j}}}\\
\vsum{V_{j}^{\prime I}(i - 1)} {a^{\prime}_{I_{j}I_{j}}}\\
\end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{\prime M}(i)} {a^{\prime}_{M_{j-1}D_{j}}}\\
\vsum{V_{j-1}^{\prime D}(i)} {a^{\prime}_{D_{j-1}D_{j}}}\\
\end{array} \right.\\[\gum]
\multicolumn3{l}{%
a'_{s_{j}s_{k}} = - \log a_{s_{j}s_{k}} 
\qquad
e^{\prime}_{s}(x) = - \log\frac{e_{s}(x)}{q_{x}}
\qquad
V_j^{\prime M}(i) = - V_j^{M}(i)
}
\\
\end{array}
\]

@@

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And we include the labels of the states that are
connected:\remark{\emph{Why} are the labels there?
And since they are apparently never used, why do they have strictness annotations???}
\smallverbatiminput{tprob}
In~a Plan7 hidden Markov model, as discussed above, 
there are seven transitions into each node:
\smallverbatiminput{tprobs}
In~our code, we select from this 7-tuple using the \texttt{ascore}
function:\remark{Norman: please put ascore into a file}
%\smallverbatiminput{ascore}
In MRFy, the first two arguments to \texttt{ascore} are known at compile
time, and we expect \texttt{ascore} to be inlined.

Nodes are numbered, and the representation of a node includes 
the tables
of emission probabilities for the match and insertion states,
plus
the
probabilities of transitions 
into the states of that node.
\smallverbatiminput{hmmnode}

@@

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}, 
which chooses the best (\texttt{minimum}) solution
from among the three predecessor states:
\smallverbatiminput{vfix}

In this code, we look up emission scores for a given node and amino
acid using \texttt{eScore},
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{extend} to add \texttt{Mat} to the list of states
encountered up to this point.

>>>>

\def\goo{18pt}
\def\gum{14pt}
\[
\begin{array}{l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{M}(i - 1)} {\log a_{M_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{I}(i - 1)} {\log a_{I_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{D}(i - 1)} {\log a_{D_{j-1}M_{j}}}\\
\end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j}^{M}(i - 1)} {\log a_{M_{j}I_{j}}}\\
\vsum{V_{j}^{I}(i - 1)} {\log a_{I_{j}I_{j}}}\\
\end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{M}(i)} {\log a_{M_{j-1}D_{j}}}\\
\vsum{V_{j-1}^{D}(i)} {\log a_{D_{j-1}D_{j}}}\\
\end{array} \right.\\
\end{array}
\qquad
\begin{array}{l@{}c@{}l}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{\prime M}(i - 1)} {a^{\prime}_{M_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{\prime I}(i - 1)} {a^{\prime}_{I_{j-1}M_{j}}}\\
\vsum{V_{j-1}^{\prime D}(i - 1)} {a^{\prime}_{D_{j-1}M_{j}}}\\
\end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j}^{\prime M}(i - 1)} {a^{\prime}_{M_{j}I_{j}}}\\
\vsum{V_{j}^{\prime I}(i - 1)} {a^{\prime}_{I_{j}I_{j}}}\\
\end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
\begin{array}{l@{}c@{}l}
\vsum{V_{j-1}^{\prime M}(i)} {a^{\prime}_{M_{j-1}D_{j}}}\\
\vsum{V_{j-1}^{\prime D}(i)} {a^{\prime}_{D_{j-1}D_{j}}}\\
\end{array} \right.\\[\gum]
\multicolumn3{l}{%
a'_{s \hat{s}} = - \log a_{s \hat{s}} 
\qquad
e^{\prime}_{s}(x) = - \log\frac{e_{s}(x)}{q_{x}}
\qquad
V_j^{\prime M}(i) = - V_j^{M}(i)
}
\\
\end{array}
\]

@@

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And we include the labels of the states that are
connected:\remark{\emph{Why} are the labels there?
And since they are apparently never used, why do they have strictness annotations???}
\smallverbatiminput{tprob}
In~a Plan7 hidden Markov model, as discussed above, 
there are seven transitions into each node:
\smallverbatiminput{tprobs}
In~our code, we select from this 7-tuple using the \texttt{aScore}
function;
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {{\hat s}}$},
where $s$~and~$\hat s$ form one of the 7~pairs above.

A~7-tuple of scores is included in the representation of each node in
the model.
This representation also includes tables of emission probabilities
$e_{M_j}$ and $e_{I_j}$.
These tables are read using function
\mbox{$\mathtt{eScore}\;s\;j\;i = e_{s_j}(x_i)$}.

%%  DROP THIS PARA??
%%  Nodes are numbered, and the representation of a node includes 
%%  the tables
%%  of emission probabilities for the match and insertion states,
%%  plus
%%  the
%%  probabilities of transitions 
%%  into the states of that node.
%%  \smallverbatiminput{hmmnode}

@@

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds $e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum.
The sum of an~$a'$~term and a $V'_{j-1}(i-1)$ term 
is computed by function \texttt{avSum},
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively.
\smallverbatiminput{vfix}
The entire right-hand side is wrapped in \texttt{extend~Mat}.
This call performs a computation \emph{not} shown in \figref{viterbi}:
in~addition to computing the most likely BUZZWORD,
\mrfy\ also computes most likely path (sequence of states).
Function \texttt{extend} adds~$M$ to the head of that path.
